### 2021-04-30 report/submission.tex
added codes in appendix.

@@

		A two dimensional random walk is a process that describes the path of an object in a two dimensional plane consisting of successive random steps. In physics, these processes play an important role in study of polymers, Brownian motion, diffusion, etc. In this problem, such a random walk is visualized. This problem also investigates the number distribution of particles with distance, after a given number of steps in random walks are executed.

@@

			\item 
				\subitem \textbf{Methods  --} 
					Random walk of a particle which is initially located at origin (0,0), is computed for 2000 time steps. At each time step, the object moves a distance, \(d=0.01\) units in a random direction which is mathematically represented using angle, \(\theta \in [0,2\pi]\), computed in the Matlab program with the help of inbuilt function, \texttt{rand()}. This function generates random number uniformly in the range \([0,1]\), therefore, multiplying the obtained random number by \(2\pi\) give us the random angle \(\theta\), in line 6 of the code \ref{code:1.1} Successive movements in the two dimensional plane are computed in lines 7 and 8 of code \ref{code:1.1}, which are enclosed in a \emph{for loop} defined in line 5, which repeats the process for each of the 2000 time steps.

@@

			\item 
				\subitem \textbf{Methods  --} 
					Random walk for 2000 time steps, like the one performed in previous part is performed for 2000 different particles. Accordingly, the crux of code \ref{code:1.1} is executed in a \emph{for loop}, 2000 times, i.e. once for each particle, as shown in line 4 of code \ref{code:1.2}. Instead of visualizing the complete path of random walk, this time only the final coordinates of the particle after finishing the random walk are considered and stored separately in lines 12 and 13 of code \ref{code:1.2}.

@@

			\item 
				\subitem \textbf{Methods  --} 
					Using elements from code \ref{code:1.2}, final positions of \(5 \times 10^{5}\) particles are computed. The functions \texttt{tic} and \texttt{toc} are used to compute the time elapsed in execution of the code. With the help of these functions, the total number of particles is chosen, based purely on estimate of the maximum computations that can be performed in feasible time. So, The initial 18 lines of code \ref{code:1.3} are essentially adopted from code \ref{code:1.2}, with an exception of line 17, where the modulus of displacement of a particle from origin is calculated. The following lines evaluate and plot the histogram of number distribution of particles with distance travelled. Using a bin width of \(dr=0.05\), binedges are computed and the histogram is plotted on line 24.\\
					An alternate method in which counts in each bin are calculated to produce the histogram, was used to confirm results, and is not shown in code \ref{code:1.3}.

@@

			\item 
				\subitem \textbf{Methods  --} 
					Initial lines of code \ref{code:1.4}, till line 27, are adopted from code \ref{code:1.3} and the value of \texttt{np} is set to \(10^5\), i.e. histogram is generated for \(10^5\) particles. Theoretically expected number distribution is then computed for each bin in lines 30-33, and the obtained curve is overlayed on the histogram. Analytically calculated number distribution of particles with distance is given by equation (3) of project notes as,
					\begin{equation}
						N(r,r+\Delta r) = N^{}_{p} \exp\left( -\frac{r^2_{}}{nd^2_{}} \right) \left\lbrace 1- \exp \left( - \frac{\Delta r (2r+\Delta r)}{nd^2_{}} \right) \right\rbrace ,
						\label{eq:number_distribution}
					\end{equation}
					where, \(N(r,r+\Delta r)\) is the number of particles that have eventually travelled an absolute value of displacement in the range \([r,r+\Delta r]\), \(N_p^{}\) is the total number of particles, \(n\) is the number of time steps and \(d\) is the jump size at each time step. This equation \ref{eq:number_distribution} is encoded in code \ref{code:1.4} in lines 31 and 32.
					\begin{figure} [h]
						\lstinputlisting[caption=(Part of) prob1ad.m - Analytical estimates and observed 	number distributions with distance of \(10^5\) particles, 
										label=code:1.4, firstline=28, firstnumber=28,
										lastline=35]{../problem1/prob1ad.m}
					\end{figure}

@@

			\item
				\subitem \textbf{Methods --}
					The number distribution of photons in this case is given by,
					\begin{equation}
						N(r) dr = 2 \pi r \rho (r) dr,
						\label{eq:photon_num_distribution}
					\end{equation}
					where, \(\rho (r)\) is the particle density distribution, given by,
					\begin{equation}
						\rho (r) = \frac{N_p^{}}{\pi n d^2_{}} \exp \left( - \frac{r^2_{}}{nd^2_{}} \right)
						\label{eq:photon_particle_density} 
					\end{equation}
					Substituting equation \ref{eq:photon_particle_density} in equation \ref{eq:photon_num_distribution}, we obtain,
					\[N(r)dr = \frac{2rN_p^{}}{nd^2_{}} \exp \left( - \frac{r^2_{}}{nd^2_{}} \right) dr.\]
					At peak number density, derivative of \(n\) with respect to displacement \(r\) equals zero. From this we obtain,
					\[\frac{dN}{dr} = \frac{2N_p^{}}{nd^2_{}} \frac{d}{dr} \left[ r \exp \left( - \frac{r^2_{}}{nd^2_{}} \right) \right] = 0 \]
					\[\Rightarrow \frac{2N_p}{nd^2_{}} \exp \left( - \frac{r^2_{}}{nd^2_{}} \right) \left[ 1- \frac{2r^2_{}}{nd^2_{}} \right] = 0 \]

@@

		For a large sample of random walk data, the number distribution of particles with eventual distance is given by equation \ref{eq:number_distribution}. The theoretical estimates and observations made using numerical simulation seem to agree well. One example of an approximate random walk is the motion of a photon through the layers of Sun. This causes the photon to take several years (of the order of millions) to reach the surface of the Sun before eventually escaping.

@@

%	\clearpage
	\setcounter{section}{0}
	\part{Circular Binary} \label{Problem2}

@@

		\section{Introduction} \label{2:introduction}
		A system of two stars rotating in circular orbits around their common centre of mass is called a \emph{circular binary}. The heavier star is called the primary star whereas the lighter one is called the secondary star. The orbital time period for both the stars is equal and therefore, primary star traverses a small orbit at slower velocity, while the orbit of the secondary star is larger which it traverses at a higher velocity. In most of the commonly encountered systems, evolutionary correlation between the two stars is extremely significant. Such a gravitationally bound system possesses a negative total energy.

@@

			\item
				\subitem \textbf{Methods  --}
				The dimensionless masses and positions of primary and secondary stars are respectively given by,
				\begin{equation}
					\begin{gathered}
						\tilde{m}_p^{}=\frac{m_p^{}}{m} \text{ and } \tilde{m}_s^{}=\frac{m_s^{}}{m}, \\
						\tilde{x}_p^{}=\frac{x_p^{}}{a}, \text{ and } \tilde{x}_s^{}=\frac{x_s^{}}{a},
					\end{gathered}
					\label{eq:dimensionless_mass_pos}
				\end{equation}
				where, \(m\) and \(a\) are the total mass and separation of binary respectively. For a circular binary, the gravitational force experienced by each star equals the centripetal force on the star. Mathematically,
				\begin{equation}
					\begin{gathered}
						\frac{Gm_{s}^{}m_{p}^{}}{a^2_{}} = m_p^{} x_p^{} \omega^2_{} = m_s^{} x_s^{} \omega^2_{} , \\
						\Rightarrow x_s^{} = \frac{x_p^{} m_p^{}}{m_s^{}}.
					\end{gathered}
					\label{eq:secondary_pos}
				\end{equation}
				Total separation of binary is the sum of positions of primary and secondary stars with respect to their centre of mass, \textit{i.e.} 
				\begin{equation}
					x_p^{}+x_s^{} = a.
					\label{eq:binary_separation}
				\end{equation}
				Using equations \ref{eq:secondary_pos} and \ref{eq:binary_separation}, we obtain,
				\begin{equation}
					\begin{gathered}
						x_p + \frac{x_p m_p}{m_s} = a \\
						\Rightarrow x_p \left( \frac{m}{m_s}\right) = a \\
						\Rightarrow \tilde{x}_p = \tilde{m}_s 
					\end{gathered}
					\label{eq:xp_tilde}
				\end{equation}
				This relation gives us the relation for dimensionless position of primary star. Similarly, we can also obtain the relation for dimensionless position of secondary star as,
				\begin{equation}
					\begin{gathered}
						x_s + \frac{x_s m_s}{m_p} = a \\
						\Rightarrow \tilde{x}_s = \tilde{m}_p
					\end{gathered}
					\label{eq:xs_tilde}
				\end{equation}

@@

				\subitem \textbf{Results  --}
				In the given scenario, the stars lie in x-y plane, with their centre of mass at origin. Primary star has initial coordinates \((-x_{px},0)\) while the secondary stars lies initially at \((x_{py},0)\).\\
				The stars rotate counter-clockwise, which gives us the velocity of primary star, \(-v_{py}\) and the velocity of secondary star, \(v_{sy}\).
				Using equations \ref{eq:xp_tilde}, \ref{eq:xs_tilde}, \ref{eq:primary_velocity} and \ref{eq:secondary_velocity}, the dimensionless values of initial conditions of the stars are summarized in table \ref{table:2_initial_conditions}.\\ The value of dimensionless time period of binary is derived in equation \ref{eq:dimensionless_time} as \(2 \pi\).

				\begin{table}
					\centering
					\begin{tabular} {c c c c}
						\toprule
						\textbf{Parameter} & \textbf{Value} & \textbf{Parameter} & \textbf{Value} \\
						(Primary Star) & & (Secondary star) & \\
						\midrule
						\(\tilde{x}_{px}\) & \(-\tilde{m}_s\) & \(\tilde{x}_{sx}\) & \(\tilde{m}_p\) \\
						\(\tilde{x}_{py}\) & 0 & \(\tilde{x}_{sy}\) & 0\\
						\(\tilde{v}_{px}\) & 0 & \(\tilde{v}_{sx}\) & 0 \\
						\(\tilde{v}_{py}\) & \(-\tilde{m}_s\) & \(\tilde{v}_{sy}\) & \(\tilde{m}_p\) \\
						\bottomrule
					\end{tabular}
				\caption{A summary of initial conditions for the simulation. All quantities are mentioned in dimensionless units, for a binary in x-y plane rotating counter-clockwise, with centre of mass at origin and primary star on the negative x-axis and secondary star on positive x-axis initially.}
				\label{table:2_initial_conditions}
				\end{table}

@@

			\item
				\subitem \textbf{Methods  --}
				Given that,
				\begin{equation}
					\frac{m_p}{m_s} = 4
					\label{eq:masses_ratio}
				\end{equation}
				Adding 1 on both side, we get,
				\begin{equation}
					\begin{gathered}
						\nonumber \frac{m_p}{m_s} + 1 = 5 \\
						\Rightarrow \frac{m_p+m_s}{m_s} = \frac{m}{m_s} = 5
					\end{gathered}
				\end{equation}
				\begin{equation}
					\Rightarrow \tilde{m}_s = \frac{m_s}{m} = \frac{1}{5} = 0.2
					\label{eq:secondary_dimensionless_mass}
				\end{equation}

				Using equations \ref{eq:dimensionless_mass_pos}, \ref{eq:masses_ratio} and \ref{eq:secondary_dimensionless_mass}, we can obtain the dimensionless mass of primary star as,
				\begin{equation}
					\tilde{m_p} = 4 \times \tilde{m_s} = 0.8
					\label{eq:primary_dimensionless_mass}
				\end{equation}

				\subitem \textbf{Results  --}
				Using relations mentioned in table \ref{table:2_initial_conditions}, the initial conditions for the circular binary are defined in Matlab file \textit{binary.m} a shown in lines 13--20 of code \ref{code:2.1}. Using equations \ref{eq:secondary_dimensionless_mass} and \ref{eq:primary_dimensionless_mass}, the values of dimensionless masses of two stars are assigned in lines 11 and 12 as \texttt{mp = 0.8} and \texttt{ms = 0.2}.\\				
				The value of \texttt{tmax} is set ot \(10\pi\). Since from equation \ref{eq:dimensionless_time}, it is known that the value of dimensionless time is \(2\pi\), this given value of \texttt{tmax} will allow simulation to sun for 5 orbits of the binary.

@@

			\begin{figure} [h]
				\begin{subfigure} {.475\columnwidth}
					\includegraphics[width=\columnwidth]{../plots/2c_orbits.png}
					\caption{Trajectories of primary and secondary stars in a circular binary system over the first five rotations.}
					\label{fig:2c}
				\end{subfigure}
				\hfill
				\begin{subfigure} {.485\columnwidth}
					\includegraphics[width=\columnwidth]{../plots/2d_energies.png}
					\caption{Analytically estimated and numerically calculated evolution of orbital energy over five rotations. The lines are seen to overlap, indicating good agreement of theoretical expectations with numerical results.}
					\label{fig:2d}
				\end{subfigure}
				\caption{Plotted results for problem 2.}
				\label{fig:2}
			\end{figure}

@@

			\item
				\subitem \textbf{Methods  --}
				Executing the Matlab code resulted in creation of file named \emph{out}. This file is then used to plot the orbits of the two stars using the pre-existing program \emph{orbitplot.m}.

@@

				\begin{figure}
					\lstinputlisting[caption=energyplot.m - Program to plot the time evolution of analytically and numerically calculated energies of a circular binary system., label=code:2.2,
					firstline=4]{../problem2/energyplot.m}
				\end{figure}

@@

		\section{Conclusions} \label{2:conclusions}
		Using fourth order Runge-Kutta method, first order ordinary differential equations are integrated to obtain values of positions and velocities of stars in a binary star system at every time step. The total energy of the bound system is constant and negative.

@@

%	\clearpage
	\setcounter{section}{0}
	\part{Hypervelocity Stars} \label{Problem3}

@@

		Hypervelocity stars are obtained when a binary interacts tidally with a black hole. Depending on several factors, one or both stars of a binary may be ejected at extremely high velocities of the order of several thousand kilometres per second. Although another possible phenomena that might result in ejection of hypervelocity stars from a binary when one star goes supernova has been highlighted in many recent texts, this shall not be the topic of study of this problem. Here, the ejection of hypervelocity stars from binary on close encounter with a black hole and its dependence on several parameters like penetration factor, orientation of rotation, masses of stars and black hole, etc are studied.

@@

				\item
					\subitem \textbf{Methods  --}
					The function \texttt{dxdy} defined in Matlab code file \emph{f.m} is modified to include black hole gravity terms in lines 7,8,11 and 12 of code \ref{code:3.f}. These modifications are done in accordance with the equations,
					\begin{equation}
						\begin{gathered}
							E_p = \frac{1}{2} m_p v_p^2 - \frac{G m_p M}{r_p} - \frac{G m_p m_s}{r_{ps}}, \\
							E_s = \frac{1}{2} m_s v_s^2 - \frac{G m_s M}{r_s} - \frac{G m_p m_s}{r_{ps}},
						\end{gathered}
					\end{equation}
					where, \(M\) is the mass of black hole and \(r_p\) and \(r_s\) are the distances from black hole to primary and secondary stars, respectively.

					\begin{figure} [h]
						\lstinputlisting[caption=f.m - Function defining values of velocities and accelerations of the two stars along the two axes., label=code:3.f]{../problem3d/f.m}
					\end{figure}

@@

					The value of \(X_{CM}^{}\) in equation \ref{eq:3-primary_position} is given by, 
					\begin{equation}
						X_{CM}^{} = (R \cos(f), R \sin(f)),
						\label{eq:3-xcm}
					\end{equation}
					where,
					\begin{equation}
						R = \frac{2 R_p}{1+ \cos(f)}.
						\label{eq:3-r}
					\end{equation}
					The angle \(f\), called \emph{true anomaly}, is the angle from point of closest approach and is a function of time. \(R_p\) is the \emph{periastron}, which is the distance of closest approach. The initial value of true anomaly is given as,
					\begin{equation}
						f_0 = - \cos^{-1} \left( \frac{D}{5} -1 \right),
						\label{eq:3-true_anomaly} 
					\end{equation}
					where, 
					\begin{equation}
						D = \frac{R_p}{R_t}
						\label{eq:3-d}
					\end{equation}

@@

					The binary star system can be assumed to be in the similar initial conditions as the one considered in part \ref{Problem2} of this project, \textit{i.e.} a circular binary rotating counter-clockwise in x-y plane with stars initially located on x-axis (Changing the phase angle \(\phi\) updates this condition accordingly. For example, \(\phi = \pi/2\) moves the initial positions to lie on y-axis). This will give us the derivative of equation \ref{eq:3-rp}, \(\dot{r}_p\) as,
					\begin{equation}
						\begin{gathered}
							\begin{aligned}
								\dot{r}_{px} &= -m_s \sin(\phi + \pi), \\
								\dot{r}_{py} &= m_s \cos(\phi + \pi).
							\end{aligned}
						\end{gathered} 
						\label{eq:3-vp_binary}
					\end{equation}

@@

					At \(t=0\), the binary passes through the periastron. This implies that the evolution of code needs to begin at time \(t<0\), which can be obtained as a function of initial true anomaly \(f_0\) discussed in equation \ref{eq:3-true_anomaly}. This gives us the initial time of 
					\begin{equation}
						t_0 = \frac{\sqrt{2}}{3} D^{3/2} \tan \left( \frac{f_0}{2} \right) \left( 3+ \tan^2 \left( \frac{f_0}{2} \right) \right) 
						\label{eq:3-initial_time}
					\end{equation}

>>>>

\usepackage[toc,page]{appendix}
\usepackage{placeins}

@@

		A two dimensional random walk is a process that describes the path of an object in a two dimensional plane consisting of successive random steps. In physics, these processes play an important role in study of polymers, Brownian motion, diffusion, etc. In this problem, such a random walk is visualized and certain aspects of it are analysed. This problem also investigates the number distribution of particles with distance, after a given number of steps in random walks are executed.

@@

			\item 
				\subitem \textbf{Methods  --} 
					Random walk of a particle which is initially located at origin (0,0), is computed for 2000 time steps. At each time step, the object moves a distance, \(d=0.01\) units in a random direction which is mathematically represented using angle, \(\theta \in [0,2\pi]\), computed in the Matlab program with the help of inbuilt function, \texttt{rand()}. This function generates random numbers uniformly in the range \([0,1]\), therefore, multiplying the obtained random number by \(2\pi\) give us the random angle \(\theta\), in line 6 of the code \ref{code:1.1}. Successive movements in the two dimensional plane are computed in lines 7 and 8 of code \ref{code:1.1}, which are enclosed in a \emph{for loop} defined in line 5, which repeats the process for each of the 2000 time steps.

@@

			\item 
				\subitem \textbf{Methods  --} 
					Random walk for 2000 time steps, like the one performed in previous part is performed for 2000 different particles. Accordingly, the crux of code \ref{code:1.1} is executed in a \emph{for loop}, 2000 times, i.e. once for each particle, as shown in line 4 of code \ref{code:1.2}. Instead of visualizing the complete path of random walk, this time only the final coordinates of the particle after finishing the random walk are considered and stored separately by assigning new variables in lines 12 and 13 of code \ref{code:1.2}.

@@

			\item 
				\subitem \textbf{Methods  --} 
					Using elements from code \ref{code:1.2}, final positions of \(5 \times 10^{5}\) particles are computed. The functions \texttt{tic} and \texttt{toc} are used to compute the time elapsed in execution of the code. With the help of these functions, the total number of particles is chosen, based purely on estimate of the maximum computations that can be performed in feasible time. So, the initial 18 lines of code \ref{code:1.3} are  adopted from code \ref{code:1.2}, with an exception of line 17, where the modulus of displacement of a particle from origin is calculated. Subsequent lines evaluate and plot the histogram of number distribution of particles with distance travelled. Using a bin width of \(dr=0.05\), binedges are computed and the histogram is plotted on line 24.\\
					An alternate method in which counts in each bin are calculated to produce the histogram, was used to confirm results, and is not shown in code \ref{code:1.3}.

@@

			\item 
				\subitem \textbf{Methods  --} 
					Initial lines of code \ref{code:1.4}, till line 27, are adopted from code \ref{code:1.3} and the value of \texttt{np} is set to \(10^5\), i.e. histogram is generated for \(10^5\) particles. Theoretically expected number distribution is then computed for each bin in lines 30-33, and the obtained curve is overlayed on the histogram. Analytically calculated number distribution of particles with distance is given by equation (3) of project notes as,
					\begin{equation}
						N(r,r+\Delta r) = N^{}_{p} \exp\left( -\frac{r^2_{}}{nd^2_{}} \right) \left\lbrace 1- \exp \left( - \frac{\Delta r (2r+\Delta r)}{nd^2_{}} \right) \right\rbrace ,
						\label{eq:number_distribution}
					\end{equation}
					where, \(N(r,r+\Delta r)\) is the number of particles that have eventually travelled an absolute value of displacement in the range \([r,r+\Delta r]\), \(N_p^{}\) is the total number of particles, \(n\) is the number of time steps and \(d\) is the jump size at each time step. This equation \ref{eq:number_distribution} is encoded in code \ref{code:1.4} in lines 31 and 32.
					\begin{figure} [h]
						\lstinputlisting[caption=(Snippet of) prob1ad.m - Analytical estimates and observed 	number distributions with distance of \(10^5\) particles, 
										label=code:1.4, firstline=28, firstnumber=28,
										lastline=35]{../problem1/prob1ad.m}
					\end{figure}

@@

			\item
				\subitem \textbf{Methods --}
					The number distribution of photons in this case is given by,
					\begin{equation}
						N(r) dr = 2 \pi r \rho (r) dr,
						\label{eq:photon_num_distribution}
					\end{equation}
					where, \(\rho (r)\) is the particle density distribution, given by,
					\begin{equation}
						\rho (r) = \frac{N_p^{}}{\pi n d^2_{}} \exp \left( - \frac{r^2_{}}{nd^2_{}} \right)
						\label{eq:photon_particle_density} 
					\end{equation}
					Substituting equation \ref{eq:photon_particle_density} in equation \ref{eq:photon_num_distribution}, we obtain,
					\[N(r)dr = \frac{2rN_p^{}}{nd^2_{}} \exp \left( - \frac{r^2_{}}{nd^2_{}} \right) dr.\]
					At peak number density, derivative of \(N\) with respect to displacement \(r\) equals zero. From this we obtain,
					\[\frac{dN}{dr} = \frac{2N_p^{}}{nd^2_{}} \frac{d}{dr} \left[ r \exp \left( - \frac{r^2_{}}{nd^2_{}} \right) \right] = 0 \]
					\[\Rightarrow \frac{2N_p}{nd^2_{}} \exp \left( - \frac{r^2_{}}{nd^2_{}} \right) \left[ 1- \frac{2r^2_{}}{nd^2_{}} \right] = 0 \]

@@

		For a large sample of random walks, the distribution of eventual positions of particles is given by equation \ref{eq:number_distribution}. The theoretical estimates and observations made using numerical simulation seem to agree well. One example of (approximation of) random walk considered here is the motion of a photon through the layers of Sun. This causes the photon to take several years (of the order of millions) to reach the surface of the Sun before eventually escaping.

@@

	\clearpage
	\setcounter{section}{0}
	\part{Circular Binary} \label{Problem2}

@@

		\section{Introduction} \label{2:introduction}
		A system of two stars rotating in circular orbits around their common centre of mass is called a \emph{circular binary}. The heavier star is called the primary star whereas the lighter one is called the secondary star. The orbital time period for both the stars is equal and therefore, primary star traverses a small orbit at slower velocity, while the orbit of the secondary star is larger which it traverses at a higher velocity. In most of the commonly encountered systems, correlation between the two stars is extremely significant in terms of their evolution. Such a gravitationally bound system possesses a negative total energy.

@@

			\item
				\subitem \textbf{Methods  --}
				The dimensionless masses and positions of primary and secondary stars are respectively given by,
				\begin{equation}
					\begin{gathered}
						\tilde{m}_p^{}=\frac{m_p^{}}{m} \text{ and } \tilde{m}_s^{}=\frac{m_s^{}}{m}, \\
						\tilde{x}_p^{}=\frac{x_p^{}}{a}, \text{ and } \tilde{x}_s^{}=\frac{x_s^{}}{a},
					\end{gathered}
					\label{eq:dimensionless_mass_pos}
				\end{equation}
				where, \(m\) and \(a\) are the total mass and separation of binary respectively. For a circular binary, the gravitational force experienced by each star equals the centripetal force on the star. Mathematically,
				\begin{equation}
					\begin{gathered}
						\frac{Gm_{s}^{}m_{p}^{}}{a^2_{}} = m_p^{} x_p^{} \omega^2_{} = m_s^{} x_s^{} \omega^2_{} , \\
						\Rightarrow x_s^{} = \frac{x_p^{} m_p^{}}{m_s^{}}.
					\end{gathered}
					\label{eq:secondary_pos}
				\end{equation}
				Total separation of binary is the sum of positions of primary and secondary stars with respect to their centre of mass, \textit{i.e.} 
				\begin{equation}
					x_p^{}+x_s^{} = a.
					\label{eq:binary_separation}
				\end{equation}
				Using equations \ref{eq:secondary_pos} and \ref{eq:binary_separation}, we obtain,
				\begin{equation}
					\begin{gathered}
						x_p + \frac{x_p m_p}{m_s} = a \\
						\Rightarrow x_p \left( \frac{m}{m_s}\right) = a \\
						\Rightarrow \tilde{x}_p = \tilde{m}_s 
					\end{gathered}
					\label{eq:xp_tilde}
				\end{equation}
				This equation gives us the relation between dimensionless position of primary star and dimensionless mass of secondary star. Similarly, we can also obtain the relation for dimensionless position of secondary star as,
				\begin{equation}
					\begin{gathered}
						x_s + \frac{x_s m_s}{m_p} = a \\
						\Rightarrow \tilde{x}_s = \tilde{m}_p
					\end{gathered}
					\label{eq:xs_tilde}
				\end{equation}

@@

				\subitem \textbf{Results  --}
				In the given scenario, the stars lie in x-y plane, with their centre of mass at origin. Primary star has initial coordinates \((-x_{px},0)\) while the secondary stars lies initially at \((x_{py},0)\).\\
				The stars rotate counter-clockwise, which gives us the velocity of primary star, \(-v_{py}\) and the velocity of secondary star, \(v_{sy}\).
				Using equations \ref{eq:xp_tilde}, \ref{eq:xs_tilde}, \ref{eq:primary_velocity} and \ref{eq:secondary_velocity}, the dimensionless values of initial conditions of the stars are summarized in table \ref{table:2_initial_conditions}.

				The value of dimensionless time period of binary is derived in equation \ref{eq:dimensionless_time} as \(2 \pi\).\\

				\begin{table} [h]
					\centering
					\begin{tabular} {c c c c}
						\toprule
						\textbf{Parameter} & \textbf{Value} & \textbf{Parameter} & \textbf{Value} \\
						(Primary Star) & & (Secondary star) & \\
						\midrule
						\(\tilde{x}_{px}\) & \(-\tilde{m}_s\) & \(\tilde{x}_{sx}\) & \(\tilde{m}_p\) \\
						\(\tilde{x}_{py}\) & 0 & \(\tilde{x}_{sy}\) & 0\\
						\(\tilde{v}_{px}\) & 0 & \(\tilde{v}_{sx}\) & 0 \\
						\(\tilde{v}_{py}\) & \(-\tilde{m}_s\) & \(\tilde{v}_{sy}\) & \(\tilde{m}_p\) \\
						\bottomrule
					\end{tabular}
				\caption{A summary of initial conditions for the simulation. All quantities are mentioned in dimensionless units, for a binary in x-y plane rotating counter-clockwise, with centre of mass at origin and primary star on the negative x-axis and secondary star on positive x-axis initially.}
				\label{table:2_initial_conditions}
				\end{table}

@@

			% ---------------------------------- (b) ---------------------------------------------
			\newpage
			\item
				\subitem \textbf{Methods  --}
				Given that,
				\begin{equation}
					\frac{m_p}{m_s} = 4
					\label{eq:masses_ratio}
				\end{equation}
				Adding 1 on both side, we get,
				\begin{equation}
					\begin{gathered}
						\nonumber \frac{m_p}{m_s} + 1 = 5 \\
						\Rightarrow \frac{m_p+m_s}{m_s} = \frac{m}{m_s} = 5
					\end{gathered}
				\end{equation}
				\begin{equation}
					\Rightarrow \tilde{m}_s = \frac{m_s}{m} = \frac{1}{5} = 0.2
					\label{eq:secondary_dimensionless_mass}
				\end{equation}

				Similarly, using equations \ref{eq:dimensionless_mass_pos}, \ref{eq:masses_ratio} and \ref{eq:secondary_dimensionless_mass}, we can obtain the dimensionless mass of primary star as,
				\begin{equation}
					\tilde{m_p} = 4 \times \tilde{m_s} = 0.8
					\label{eq:primary_dimensionless_mass}
				\end{equation}

				\subitem \textbf{Results  --}
				Using relations mentioned in table \ref{table:2_initial_conditions}, the initial conditions for the circular binary are defined in Matlab file \textit{binary.m} a shown in lines 13--20 of code \ref{code:2.1}. Using equations \ref{eq:secondary_dimensionless_mass} and \ref{eq:primary_dimensionless_mass}, the values of dimensionless masses of two stars are assigned in lines 11 and 12 as \texttt{mp = 0.8} and \texttt{ms = 0.2}.\\				
				The value of \texttt{tmax} is set to \(10\pi\). Since from equation \ref{eq:dimensionless_time}, it is known that the value of dimensionless time is \(2\pi\), this given value of \texttt{tmax} will allow simulation to sun for 5 orbits of the binary.

@@

			\begin{figure} [h]
				\begin{subfigure} {.475\columnwidth}
					\includegraphics[width=\columnwidth]{../plots/2c_orbits.png}
					\caption{Trajectories of primary and secondary stars in a circular binary system over the first five rotations.}
					\label{fig:2c}
				\end{subfigure}
				\hfill
				\begin{subfigure} {.485\columnwidth}
					\includegraphics[width=\columnwidth]{../plots/2d_energies.png}
					\caption{Analytically estimated and numerically calculated evolution of orbital energy over five rotations. The lines are seen to overlap, indicating good agreement of numerical results with theoretical expectations.}
					\label{fig:2d}
				\end{subfigure}
				\caption{Plotted results for problem 2.}
				\label{fig:2}
			\end{figure}

@@

			\item
				\subitem \textbf{Methods  --}
				Executing the Matlab code resulted in creation of file named \emph{out}. Data obtained in this file is then used to plot the orbits of the two stars using the pre-existing program \emph{orbitplot.m}.

@@

				\begin{figure} [h]
					\lstinputlisting[caption=energyplot.m - Program to plot the time evolution of analytically and numerically calculated energies of a circular binary system., label=code:2.2,
					firstline=4]{../problem2/energyplot.m}
				\end{figure}

@@

		\section{Conclusions} \label{2:conclusions}
		This initial positions and velocity components of primary and secondary stars in a circular binary have been calculated. Using fourth order Runge-Kutta method, first order ordinary differential equations are integrated to obtain values of positions and velocities of stars in a binary star system at subsequent time steps. The total energy of the bound system is constant and negative.

@@

	\clearpage
	\setcounter{section}{0}
	\part{Hypervelocity Stars} \label{Problem3}

@@

		Hypervelocity stars are obtained when a binary star system interacts tidally with a black hole. Depending on several factors, one or both stars of a binary may be ejected at extremely high velocities of the order of several thousand kilometres per second. Although another possible phenomenon that might result in ejection of hypervelocity stars from a binary when one star goes supernova has been highlighted in many recent texts, this shall not be the topic of study of this problem. Here, the ejection of hypervelocity stars from binary on close encounter with a black hole and its dependence on several parameters like penetration factor, orientation of rotation, masses of stars and black hole, etc. are studied.

@@

				\item
					\subitem \textbf{Methods  --}
					The function \texttt{dxdy} defined in Matlab code file \emph{f.m} is modified to include black hole gravity terms in lines 7,8,11 and 12 of code \ref{code:3.f}. These modifications are done according to the equations,
					\begin{equation}
						\begin{gathered}
							E_p = \frac{1}{2} m_p v_p^2 - \frac{G m_p M}{r_p} - \frac{G m_p m_s}{r_{ps}}, \\
							E_s = \frac{1}{2} m_s v_s^2 - \frac{G m_s M}{r_s} - \frac{G m_p m_s}{r_{ps}},
						\end{gathered}
					\end{equation}
					where, \(M\) is the mass of black hole and \(r_p\) and \(r_s\) are the distances from black hole to primary and secondary stars, respectively.

					\begin{figure}
						\lstinputlisting[caption=f.m - Function defining values of velocities and accelerations of the two stars along the two axes., label=code:3.f]{../problem3d/f.m}
					\end{figure}

@@

					The value of \(X_{CM}^{}\) in equation \ref{eq:3-primary_position} is given by, 
					\begin{equation}
						X_{CM}^{} = (R \cos(f), R \sin(f)),
						\label{eq:3-xcm}
					\end{equation}
					where,
					\begin{equation}
						R = \frac{2 R_p}{1+ \cos(f)}.
						\label{eq:3-r}
					\end{equation}
					The angle \(f\), called \emph{true anomaly}, is the angle from point of closest approach and is a function of time. \(R_p\) is the \emph{periastron}, which is the distance of closest approach. The initial value of true anomaly is given as,
					\begin{equation}
						f_0 = - \cos^{-1} \left( \frac{D}{5} -1 \right),
						\label{eq:3-true_anomaly} 
					\end{equation}
					where, 
					\begin{equation}
						D = \frac{R_p}{R_t},
						\label{eq:3-d}
					\end{equation}
					is called the penetration factor.\\

@@

					The binary star system can be assumed to be in the similar initial conditions as the one considered in part \ref{Problem2} of this project, \textit{i.e.} a circular binary rotating counter-clockwise in x-y plane with stars initially located on x-axis (Changing the phase angle \(\phi\) updates this condition accordingly). This will give us the derivative of equation \ref{eq:3-rp}, \(\dot{r}_p\) as,
					\begin{equation}
						\begin{gathered}
							\begin{aligned}
								\dot{r}_{px} &= -m_s \sin(\phi + \pi), \\
								\dot{r}_{py} &= m_s \cos(\phi + \pi).
							\end{aligned}
						\end{gathered} 
						\label{eq:3-vp_binary}
					\end{equation}

@@

					At \(t=0\), the binary passes through the periastron. This implies that the evolution of code needs to begin at time \(t<0\), which can be obtained as a function of initial true anomaly \(f_0\) discussed in equation \ref{eq:3-true_anomaly}. This gives us the initial time of 
					\begin{equation}
						t_0 = \frac{\sqrt{2}}{3} D^{3/2} \tan \left( \frac{f_0}{2} \right) \left( 3+ \tan^2 \left( \frac{f_0}{2} \right) \right) .
						\label{eq:3-initial_time}
					\end{equation}

					This gives us a complete set of initial conditions that can be used for simulation -- initial time, initial positions and initial velocities. These relations can be used in the Matlab file \emph{initialc.m} given in code \ref{code:3.initialc}, to define the initial conditions. \\
